Historical collections as baseline for assessing the extent of the pollinator crisis

### 2018-04-01 ms.Rmd
Draft in quite good shape

---
title: "Museum specimens as baseline for assessing a potential global pollinator crisis"
author: "I. Bartomeus, J. Stavert, D. Ward, O. Aguado"
csl: data/philb_style.xml
output:
  html_document: default
  word_document: default
bibliography: data/references.xml
---

@@

@@

Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [#@Ollerton2012] and food security [@Klein2007]. As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions. The vulnerability of pollinators was identified several decades ago, and was popularized in 1996 by the influential book “the forgotten pollinators” [@Buchman1996]. However, early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time. These initial claims triggered the first efforts to assess the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines. The NAS report concluded that “For most pollinator species […] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult” [@nas]. Since then, studies on pollinator responses to various global change drivers have multiplied rapidly. Researchers have now developed strong consensus that disturbances such as habitat destruction, land-use intensification, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010]. Yet, the current status and population trends of most pollinator species worldwide remain unknown. For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species check number fell into the “Data Deficient” category [@Nieto2014]. For countries outside of Europe and the US, data on pollinator populations is almost non-existent.

One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and include bees, flies, butterflies, beetles, birds, bats and even lizards [#@Winfree2011]. Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017]. However, bees are also diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and often require expert taxonomists for identification. Furthermore, the uneven distribution of pollinator researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos, @DePalma2016]. Another barrier to assessing pollinator declines is that pollinators are often highly mobile, short-lived and small.

One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients. Despite critiques on the robustness of this approach [@Franca, @DePalma2018], this is the most extensive source of data on pollinator populations currently available. In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models using pollinator associations to different land uses [@DePalma2017]. A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists. This approach was inspired by successful butterfly monitoring programs [@vanSwaay] and is currently being extended to other pollinator taxa. However, these programs require significant time to generate long-term datasets and cannot reveal anything about historic pollinator populations. Finally, the most practical approach for assessing long-term pollinator population trends is to use museum specimen collection data.

The core aim of museums is to conserve and curate collections. Thus, they serve as a precious repository for specimens, and at the same time often ensure higher quality taxonomic identification. Yet, the major bottleneck for researchers wanting to use these data is the lack of digitization. Where digitization has been completed, the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g., [@Bartomeus2013a]; # @Scheper2014, @Young). This is despite the fact that museum collections often have a number of biases including: the unknown sampling effort that was employed; the personal interests and curatorial techniques of collectors (e.g. discarding damaged individuals, only accessioning a certain number of individuals, targeting rare or unusual over common taxa); the spatial biases where areas have been undersampled, or where samples are biased towards easily collected localities (e.g. near towns/cities and/or along roadsides);that information is often restricted only to the presence of a species (i.e. there is no information on where a species is absent);and the difficulty of getting information on other taxa from the same location (e.g. NHCs are organised taxonomically, not geographically [#@Keil; other @refs?]. Nevertheless, in many cases, museum collections are the only data we have to reconstruct the past.

In this review we first highlight the major technical difficulties involved in using museum collection data. We then assess current evidence for diversity declines and highlight the need to move beyond simple diversity descriptors [this is not very clear – are we talking about richness?]. Finally we present a road map outlining a strategy for using museum data to fill current knowledge gaps. We demonstrate how to approach analysis of different types for museum collection data to identify long-term trends in pollinator populations.

#The use of museum specimens

Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@taxonomy], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@hart]. Undertaking this process for tens or hundreds of thousands of museum collections specimens can be a daunting task and requires specialized personnel. While some tasks can only be undertaken by people with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the collection digitization process. High resolution photos of specimens and associated labels can be uploaded to the internet, where the job of image transcription can be distributed across hundreds or thousands of volunteers. In addition, new algorithms have been created that allow location geo-referencing based on vernacular names [@ref geonames, geoparser and opencage]. [Explain zooniverse, NZ initiative and how data can be validated… Is there anything on AI to automate the process? e.g. there is something on wing venation from photos at Netherlands. Can be mentioned?]

@@

More detailed patterns for common bee species in the Netherlands have shown simultaneous plant and pollinator declines [#@Scheper2014]. Specifically, bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus, were most threatened with extinction. [check this] [- EU: UK? Ollerton Science, Peter Keil flies? Moths? Butterflies! Add Ollerton paper!]. Although studies of local pollinator communities often detect richness declines, regional richness may remain relatively stable. For example, regional estimates for species richness changes in the eastern US show moderate declines [@Bartomeus2013a] and very few regional extinctions [@Colla2012]. These findings are in stark contrast with the widespread local extinctions reported by Burkle [@Burkle2013], who compared historical observations of bee occurrence in a large forested ecosystem with remaining forest remnants. However, it is important to note that this study focused on habitat remnants and did not take into account opportunistic bee species that use novel land-use types (#@collado2018). In any case, there is strong concordance between local extinctions and regional declines [#@Bartomeus_Winfree2013], suggesting that local extinctions are indicators of regional population declines.

@@

Outside of Europe and the US there are no studies on pollinator declines using historical museum collection records. South Africa: @Pauw, maybe with plants… . However, there are species-specific examples of historical (e.g., Bombus dalbhomi; #@Aizen).
add % in text also
Justify focus on bees? We compiled all this information to show several worldwide maps showing knowledge about bee richness (a), Bee species richness changes (b), syrphid species richness changes (c) and butterfly species richness changes to highlight the little coverage that we have globally.

```{r map, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(rworldmap)
map <- read.csv("data/bee_richness.csv")
#Jamie, if you want, we could complete this data set based on the papers.
# If two estimates available for the same country, I will use two rows. 
# We can compute the mean or the max using reshape here.
# For studies reporting large regions within country this is ok, I think, but
# for studies reporting locations (like Burkle Science paper) the % richness loss would be inflated when scaling up... not sure how to add those up. Add a column maybe saying coverage
# country, region, individual sites.
#Bee richness can be gathered from Discover life webpage
#head(map)
map2 <- joinCountryData2Map(map, joinCode="NAME", nameJoinColumn = "country",
                            nameCountryColumn = "distribution")
#need to use ISO's to make sure we don't carry over NA's due to country misspellings
#have plotted countries with missing values as black (shows we have richness for pretty much everywhere)
map3 <- mapCountryData(map2, nameColumnToPlot = "species", missingCountryCol = "light grey",
               mapTitle = "Bee species richness", numCats = 25, catMethod = "pretty", addLegend = FALSE)
#change it to grey, blak was too contrasting!

#add a modified legend using the same initial parameters as mapCountryData
do.call(addMapLegend, c(map3, legendLabels="all", legendWidth=0.5))

#Nacho needs to correct for country size

Figure 1: Global map showing bee species richness. Countries without available data are coloured in grey. Data were downloaded from the Discover Life website.

```{r}
#plot for species declines??

#Plots look mainly white... I wonder if they are useful...

[need integrate this two paragraphs!]()
need integrate this two paragraphs! Regardless of the method, it is important for researchers to go beyond aspects of biodiversity focused solely in species richness. A recent survey highlighted that more than 90% of the papers studying pollinator responses to land use change focused on richness and abundance descriptors. But in addition to local (alpha) diversity and regional (gamma) diversity, we need to measure changes in turnover between sites (beta diversity), abundance of key species, and evenness of communities [@Winfree2015]. Environmental changes often results in a few species as “winners” and many species becoming “losers” [@Bartomeus2013b]. Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2018]. These changes can have important flow on effects for pollination of native plant species and crops. Add something about traits of winners and losers? https://www.sciencedirect.com/science/article/pii/S0065250417300296

Museum can provide valuable information besides species occurrence records, and such information is important to capture when digitizing collections. This is specifically relevant to identify mechanisms of decline and adaptation. For example, one particularly important source of information is the date the specimen was collected, which allows the tracking of phenological advances congruent with contemporary climate change [#@Bartomeus2011]. In addition, museum labels on pollinator specimens often include information about the host plant on which the specimen was collected. This information is precious to understand past and present species interactions [#@Tylianakis2010ELE]. Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events [#@kleinj_raemakers]. Finally, museum specimens can be measured for specific attributes and this has been done to show tongue length [#@refAlaskaScience] and body size [#@Oliveira] shrinkage with climate and land use change.

@@

There is a paucity of countries for which museum data is available (Figure 1), and hence can be used as baseline for assessing pollinator population declines. While ideally, one would aim to digitize all museum records in the world, this is unlikely to happen in the near future due mainly to funding constraints. [from Darren: Not just about lack of funding, it is also about time. It could happen faster IF other things were not done but museums have competing priorities]. Here we illustrate how this can be prioritized and optimized to encourage independent researchers to assess declines in its own country.

GBIF (www.gbig.com?) is a central repository centralizing occurrence data from around the globe. Much of this data comes from museums and taxonomic collections [from Darren: Wider than just a ‘museum’ in the traditional sense; also a lot goes on at govt research institutes that have a collection (eg like NZAC), and also sometimes at Universities,], but it integrates several other sources. Alongside the popular statistical language R [#@R], GBIF can be directly queried into your computer [#@rgbif] and data availability checked for your region of interest. In Figure 2 we show the number of modern and historical bee records already available per country. While for some countries like UK and US the numbers are high in both axes, and hence, data can be potentially analyzed without further data collection efforts, most countries fall short in one or both of the axes. Hence, for countries with large number of recent records, researchers should prioritize the digitalization of old material to conduct their analysis. On the contrary, for countries with large number of historical records, re-surveys should be organized [e.g. #@cameron2013]. Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X… Also note that historical records are not always in local museums, but many UK and US museums contain large collection of exotic pollinators.

@@

Figure 2. The number of bee occurrences pre 1980 and post 1980 in GBIF. Upper right quadrat are well covered countries, upper left quadrat are countries that need resurvey of old sites. Lower right are sites which needs to dig museum records and lower left are countries hard to get data.

[we can point out % of countries with > 1000 records in either axes, for example, etc... would be cool to ggplot this and point where NZ and SP fall, for example. As well as the green and red dots in the lower right cuadrat]()

Once museum collection datasets are made available, researchers need to gauge any potential biases in the data. We explore this process with two contrasting examples from Spain and New Zealand. In the Spanish dataset, most of the data comes from a few specific locations and was collected by the same team of people. In this case, we contacted the original collectors to define their sampling protocols. We then resurveyed the same sites using the same protocols 20 years later. Conversely, the New Zealand dataset includes a wide suite of collectors and collection locations. For this dataset, we ensured that there is no clear differences to geographical I haven’t actually looked at this as there weren’t and GPS coordinates in the data and taxonomic collection patterns through time. For example, some collectors may have targeted particular taxa, at specific locations or during certain time periods, which would represent a bias in the data. For these case studies, we provide annotated R scripts to analyse the different kinds of data (e.g. for Spain and New Zealand, as described above). Essentially, these different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories. We hope this resource will encourage researchers to analyze data for regions where there is little current information on pollinator declines.

@@

Spain provides an interesting study system because its natural habitats have been transformed extensively by humans, but land-use is not as intensive as many other European countries. In addition, Spain is a bee diversity hotspot and maintains a relatively heterogeneous landscape. Spain has already digitalized a large amount of occurrence data for both historical and recent periods (Figure 2). However, visual inspection of the data reveals that the data is clustered around a few localities. Further, historic records do not spatially match recent records, making them difficult to compare. For this dataset, most of the historical records are located around Valladolid and were collected by Enrique Asensio and his collaborators. There has been no recent sampling of bees in this area. However, we found that Enrique systematically sampled seven locations and that there was additional data available at the “Museo de Historia Nacional” and other minor collections. Digitization of these records, along with a re-survey of these locations provided an excellent dataset for a before and after comparison of bee communities at these locations. [from Darren: Has there been any changes in the land use between the two periods?]

In brief, we check sampling completeness at both time periods, and compare the species richness level of each site before 1980 and after 1980 with a paired t-test. We show that overall there is no richness loss (p > 0.05) [Should it be “no difference in species richness”]. However, raw abundance [is this what you mean by numbers?] show a steep decline… and groups like nomadas, tyreus, prosopis and teralonias are ocmpletely extirpated (appendix?). [from Darren: Any information on these groups. Are the specialists species, or more sensitive to land change?].

@@

In global terms, human colonisation of a-New Zealand was relatively recent (c. 740 y) [@Wilmshurst2008]. Before human arrival, New Zealand was predominately forested, but has since been dramatically altered by people. Early Māori settlers cleared forests by burning and more recently, European colonists cleared large tracts of remaining forests and drained low-lying wetlands for agriculture, mostly before 1900 [@Perry2014]. Therefore, human activity likely affected pollinator communities in New Zealand long before surveys and collections began. Nevertheless, we can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.

We used New Zealand bee collection records gathered from multiple sources, including university, research institute, museum and private collections. Collection records from the New Zealand Arthropod Collection (NZAC) are freely available online (https://scd.landcareresearch.co.nz/). Fly pollinator data was obtained from six participating New Zealand museums [need to list in the acknowledgments] and covers two families (Calliphoridae and Syrphidae) that contain important fly pollinators. Collections for the bee and fly datasets span over 100 years (early 1900s to late 2000s).

First, we filtered our original datasets so that data used for analyses only included unique collection events. To do this, we removed specimens collected at the same location, on the same date, and by the same collector. We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980. We accounted for differences in collection effort through binning collection records by time so that each bin had a similar number of records but a different number of years [@Bartomeus2013a]. We then estimated richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the mean species richness ±SE for each bin. Finally, we estimated the significance of change in richness using a permutation test that randomly reordered time periods and calculated the correlation between time period and species richness. Thus, reported P-values were the proportion of permutations that had higher or lower correlations compared to the correlation between richness and the actual chronological time period sequence.

To determine if the probability of finding a species in the collection changed over time, we used a general linear model with a binomial distribution and a logit link. [should we analyse natives and exotics together or separately?]. For species that showed overdispersion, we used a quasi-binomial distribution. Further, we only included species in this analysis for which we had 30 or more records. To account for differences in sampling effort between years, we weighted each year by the total number of samples collected in that year.

@@

##Changes in species relative abundances over time

We found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic) and three bee species declined in relative abundance (one native and two exotic) (Figure 3). Interestingly, the two exotic bee species that declined in relative abundance were both in the genus Bombus, intentionally introduced into New Zealand for the pollination of crops. Native bees that increased in relative abundance were mostly from the genus Leioproctus. Only two out of 14 fly species increased in relative abundance, both of which were exotic, whereas four species decreased in abundance (three native and one exotic). Native flies that decreased in relative abundance were all Syrphidae in the genus Helophilus.

@@

- Do we need some sort of separate discussion or integrate into the results section? i would say integrate! … Although this is surprising, it could simply indicate a change in the evenness of species collected over time as the method we used actually measures both richness and evenness. Yes.

#The way forward

As illustrated in our examples, unleashing the power of museum collection data to answer pressing questions is at our hands, but requires the coordinated effort of many actors. Using two case studies, we show that collaboration with both museum curators and ecologists is key to adequately understanding the data and treating it appropriately. We envision that instead of a central museum collection record repository, researchers and curators should simply aim to make digitalized data readily available and easy to share. [from Darren: I don’t think this is a good statement. It can be deleted and the discussion is still good. Saying “instead of a central museum..” is an argument against the main part of this paper (early in introduction) and also is against the discussion about collaboration]. Ultimately, centralization of regional and national museum data in existing global platforms such as GBIF would facilitate free and widespread access. However, ecological [from Darren: Adding “ecological” means the data is different from the museum records in a central repository. Ecological datasets also do not fit well into museums databases unless there are voucher specimens from the ecology project] datasets can also be stored in different webpages or database repositories (e.g. Dryad), and can be retrieved and combined with other datasets using open science tools (ropensci), providing these tools are accessible [API e.g., # @heltmint].

We also need to revolutionize the way that researchers collaborate with museums, which should start with fostering healthy bidirectional relationships. For instance, ecological researchers collect massive amounts of specimens, but these are often inappropriately vouchered, rendering them less useful for future research. To improve this, strong communication between museums and researchers is required, but this will only be achieved if there is not adequate funding, and recognition that accurate data recording and long-term preservation are critical for research [@WardTEE].

>>>>

---
title: "Historical collections as baseline for assessing the extent of the pollinator crisis"
author: "I. Bartomeus, J. Stavert, D. Ward, O. Aguado"
csl: data/philb_style.xml
output:
  html_document: default
bibliography: data/references.xml
---

@@

[Historical collections or museum collections?]()

*Affiliations:*
1 Estación Biológica de Doñana (EBD-CSIC), Avda. Américo Vespucio 26, Isla de la Cartuja, E-41092 Sevilla, Spain 
2 
3
4 Andrena Soluciones
*Correspondence: nacho.bartomeus@gmail.com

*Author contributions*: IB wrote the initial draft. DW and OA provided data. IB and JS analyzed the data. All authors contributed to writting the manuscript.

@@

Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [#@Ollerton2012] and food security [@Klein2007]. As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions. The vulnerability of pollinators was identified several decades ago, and was popularized in 1996 by the influential book “the forgotten pollinators” [@Buchman1996]. However, early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time. These initial claims triggered the first efforts to assess the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines. The NAS report concluded that “For most pollinator species […] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult” [@nas]. Since then, studies on pollinator responses to various global change drivers have multiplied rapidly. Researchers have now developed strong consensus that disturbances such as habitat destruction, land-use intensification, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010]. Yet, the current status and population trends of most pollinator species worldwide remain unknown. For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, > 55% of bee species fell into the “Data Deficient” category [@Nieto2014]. For countries outside of Europe and the US, data on pollinator populations is almost non-existent.

One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and include bees, flies, butterflies, beetles, birds, bats and even lizards [#@Winfree2011]. All this groups of pollinators are often highly mobile, short-lived and small, difficulting the monitoring of its poplations. Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017]. However, bees are also diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and often require expert taxonomists for identification. Furthermore, the uneven distribution of researchers [#@Martinetal2012] has also resulted in geographical biases in bee decline research, as well as taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos, @DePalma2016].

One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients. Despite critiques on the robustness of this approach [@Franca, @DePalma2018], this is the most extensive source of data on pollinator populations currently available. For example, in Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models using pollinator associations to different land uses [@DePalma2017]. A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists. This approach was inspired by successful butterfly monitoring programs [@vanSwaay] and is currently being extended to other pollinator taxa. However, these programs require significant time to generate long-term datasets and cannot reveal anything about historic pollinator populations. Finally, the most practical approach for assessing long-term pollinator population trends is to use the available historical information on species occurrences deposited in museum specimen collection data [e.g. @bartomeus2013].

In this review we first assess current evidence for diversity declines and present a road map outlining a strategy for using museum data to fill current knowledge gaps. We highlight the major technical difficulties involved in using museum collection data and demonstrate how to approach analysis of different types for museum collection data to identify long-term trends in pollinator populations. Finaly, we highlight the need to move beyond simple diversity descriptors and unleash the power of historical data to assess also species interactions, ecosyetm functioning and evolutionary changes along time.

@@

More detailed patterns for common bee species in the Netherlands, aslo using museum historical specimens, have shown simultaneous plant and pollinator declines [#@Scheper2014]. Specifically, bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus, were most threatened with extinction. Although studies of local pollinator communities often detect richness declines, regional richness may remain relatively stable. For example, regional estimates for species richness changes in the eastern US show moderate declines [@Bartomeus2013a] and very few regional extinctions [@Colla2012]. A pattern also found in the UK [@OllertonScience]. The US findings are in stark contrast with the widespread local extinctions reported by Burkle [@Burkle2013], who compared historical observations of bee occurrence in a large forested ecosystem with remaining forest remnants. However, it is important to note that this study focused on habitat remnants and did not take into account opportunistic bee species that use novel land-use types (#@collado2018). In any case, there is strong concordance between local extinctions and regional declines [#@Bartomeus_Winfree2013], suggesting that local extinctions are indicators of regional population declines.

@@

Outside of Europe and the US there are no studies on pollinator declines using historical records. In Fig 1 we show a comparision of bee species richness per area (a), bee species richness changes (b), syrphid species richness changes (c) and butterfly species richness changes (d) to highlight the little coverage that we have globally. See raw data at Sup Mat 1. For other parts of the world there are only species-specific examples of historical losses (e.g., Bombus dalbhomi; #@Aizen) [South Africa: @Pauw, maybe with plants… . However,Peter Keil flies? Moths? Butterflies!]()

```{r map, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(rworldmap)
map <- read.csv("data/bee_richness.csv")
size <- read.csv("data/country_area.csv")
colnames(size) <- c("country", "size")
map0 <- merge(map, size, all.x = TRUE)
scatter.smooth(log(map0$species) ~ log(map0$size+1))
m <- lm(log(map0$species) ~ log(map0$size+1), na.action = na.exclude) 
map0$res <- residuals(m)
map2 <- joinCountryData2Map(map0, joinCode="NAME", nameJoinColumn = "country",
                            nameCountryColumn = "distribution")
#need to use ISO's to make sure we don't carry over NA's due to country misspellings
#have plotted countries with missing values as grey (shows we have richness for pretty much everywhere)
map3 <- mapCountryData(map2, nameColumnToPlot = "res", missingCountryCol = "light grey",
               mapTitle = "Bee species richness", numCats = 25, catMethod = "pretty", addLegend = FALSE)
#add a modified legend using the same initial parameters as mapCountryData
do.call(addMapLegend, c(map3, legendLabels="all", legendWidth=0.5))

#Nacho needs to correct for country size
#Need to 1) match all country names (includinh US and UK!)
#2) check subsaharian countries maybe omit?

#need to add all the pannels.
```

Figure 1: Global map showing a) bee species richness per area (Data from www.discoverlife.org). This is calculated as the residuals of the log-log regresion between bee richness and country size to account for the species-area relationships. We show mean local bee (b), syrphid (c) and butterflies (d) richness percent change in the last century. Warmer colors indicate steeper declines observed. Countries without available data are coloured in grey.

#How to use historical records vouchered in museum specimens to fill the gaps.

As evidenced in Fig 1, we lack specific estimates for most of the world countries and histrical data vouchered in museums may be the only change to fill those gaps.

The core aim of museums is to conserve and curate collections. Thus, they serve as a precious repository for specimens, and at the same time often ensure higher quality taxonomic identification. Yet, the major bottleneck for researchers wanting to use these data is the lack of digitization. Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@taxonomy, @taxonomy2, @Thomson], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@hart]. Undertaking this process for tens or hundreds of thousands of museum collections specimens can be a daunting task and requires specialized personnel. While some tasks can only be undertaken by people with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the collection digitization process. High resolution photos of specimens and associated labels can be uploaded to the internet, where the job of image transcription can be distributed across hundreds or thousands of volunteers (e.g. https://www.zooniverse.org/). In addition, new algorithms have been created that allow location geo-referencing based on vernacular names [@ref geonames, geoparser and opencage]. However, achieving all this goals needs propoer funding [@brief1].

Where digitization has been completed, the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g., [@Bartomeus2013a]; # @Scheper2014, @Young). This is despite the fact that museum collections often have a number of biases. This biases includes the unknown sampling effort that was employed or the personal interests and curatorial techniques of collectors. For example, collectors tend to target rare or unusual over common taxa, discard damaged individuals or only accession a certain number of individuals. In addition, the opportunistic nature of museum collections often imply an spatial bias where areas have been undersampled, or where samples are biased towards easily collected localities (e.g. near towns/cities and/or along roadsides) and that information is often restricted only to the presence of a species (i.e. there is no information on where a species is absent). Given adequate sample sizes, most of this biases can be accounted for with propoer statistical data anslysis [e.g. @bartomeus2013, #@Keil, @pearce]. Nevertheless, in many cases, museum collections are the only data we have to reconstruct the past.

@@

As we have seen, there is a paucity of countries for which museum data is available (Fig 1), and hence can be used as baseline for assessing pollinator population declines. While ideally, one would aim to digitize all museum records in the world, this is unlikely to happen in the near future due mainly to funding constraints. Here we illustrate how this can be prioritized and optimized to encourage independent researchers to assess declines in its own country.

GBIF (https://www.gbif.org/) is a central repository centralizing occurrence data from around the globe. Much of this data comes from museums, collections and goverment research institutes, but it integrates several other sources. Alongside the popular statistical language R [#@R], GBIF can be directly queried into your computer [#@rgbif] and data availability checked for your region of interest. In Figure 2 we show the number of modern and historical bee records already available per country. While for some countries like UK and US the numbers are high in both axes, and hence, data can be potentially analyzed without further data collection efforts, most countries fall short in one or both of the axes. Hence, for countries with large number of recent records, researchers should prioritize the digitalization of old material to conduct their analysis. On the contrary, for countries with large number of historical records, re-surveys should be organized [e.g. #@cameron2013]. Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X… Also note that historical records are not always in local museums, but many EU and US museums contain large collection of exotic pollinators. Similar exploratry analysis can be easely done for other taxa.

@@

Figure 2. The number of bee occurrences pre 1980 and post 1980 in GBIF. Upper right quadrat are well covered countries, upper left quadrat are countries that need resurvey of old sites. Lower right are sites which needs to dig museum records and lower left are countries with very little data available. We highlight a few representative examples.

Once museum collection datasets are made available, researchers need to gauge any potential biases in the data. We explore this process with two contrasting and ilustrative examples from Spain and New Zealand. In the Spanish dataset, most of the data comes from a few specific locations and was collected by the same team of people. In this case, we contacted the original collectors to define their sampling protocols. We then resurveyed the same sites using the same protocols 20 years later. Conversely, the New Zealand dataset includes a wide suite of collectors and collection locations, but luckely, tehere is no obvious bias in the geographical and taxonomic coverage through time. For these case studies, we provide annotated R scripts to analyse the different kinds of data (e.g. for Spain and New Zealand, as described above). Essentially, these different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories. We hope this resource will encourage researchers to analyze data for regions where there is little current information on pollinator declines.

@@

Spain provides an interesting study system because its natural habitats have been transformed extensively by humans, but land-use is not as intensive as in many other European countries. In addition, Spain is a bee diversity hotspot (Fig 1) and maintains a relatively heterogeneous landscape. Spain has already digitalized a large amount of occurrence data for both historical and recent periods (Fig 2). However, visual inspection of the data reveals that the data is clustered around a few localities. Further, historic records do not spatially match recent records, making them difficult to compare. For this dataset, most of the historical records are located around Valladolid and were collected by Enrique Asensio and his collaborators. There has been no recent sampling of bees in this area. However, we found that Enrique systematically sampled seven locations and that there was additional data available at the “Museo de Historia Nacional” and other minor collections. Digitization of these records, along with a re-survey of these locations provided an excellent dataset for a before and after comparison of bee communities at these locations. [from Darren: Has there been any changes in the land use between the two periods?]

In brief, we check sampling completeness at both time periods, and compare the species richness level of each site before 1980 and after 1980 with a paired t-test. We show that overall there is no no difference in species richness (p > 0.05). However, ... show a steep decline… and groups like nomadas, tyreus, prosopis and teralonias are completely extirpated (appendix?). [from Darren: Any information on these groups. Are the specialists species, or more sensitive to land change?].

@@

In global terms, human colonisation of New Zealand was relatively recent (c. 740 y) [@Wilmshurst2008]. Before human arrival, New Zealand was predominately forested, but has since been dramatically altered by people. Early Māori settlers cleared forests by burning and more recently, European colonists cleared large tracts of remaining forests and drained low-lying wetlands for agriculture, mostly before 1900 [@Perry2014]. Therefore, human activity likely affected pollinator communities in New Zealand long before surveys and collections began. Nevertheless, we can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.

We used New Zealand bee collection records gathered from multiple sources, including university, research institute, museum and private collections. Collection records from the New Zealand Arthropod Collection (NZAC) are freely available online (https://scd.landcareresearch.co.nz/). Fly pollinator data was obtained from six participating New Zealand museums [need to list in the acknowledgments]() and covers two families (Calliphoridae and Syrphidae) that contain important fly pollinators. Collections for the bee and fly datasets span over 100 years (early 1900s to late 2000s).

We follow Bartomeus et al. 2013a protocol to analysze this data. First, we filtered our original datasets so that data used for analyses only included independent collection events. To do this, we removed specimens collected at the same location, on the same date, and by the same collector. We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980. We accounted for differences in collection effort through binning collection records by time so that each bin had a similar number of records but a different number of years. We then estimated richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the mean species richness ±SE for each bin. Finally, we estimated the significance of change in richness using a permutation test that randomly reordered time periods and calculated the correlation between time period and species richness. Thus, reported P-values were the proportion of permutations that had higher or lower correlations compared to the correlation between richness and the actual chronological time period sequence.

Second, to determine if the probability of finding a species in the collection changed over time, we used a general linear model with a binomial distribution and a logit link. For species that showed overdispersion, we used a quasi-binomial distribution. Further, we only included species in this analysis for which we had 30 or more records. To account for differences in sampling effort between years, we weighted each year by the total number of samples collected in that year.

@@

However, at the species level, we found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic) and three bee species declined in relative abundance (one native and two exotic) (Fig 3). Interestingly, the two exotic bee species that declined in relative abundance were both in the genus Bombus, intentionally introduced into New Zealand for the pollination of crops. Native bees that increased in relative abundance were mostly from the genus Leioproctus. Only two out of 14 fly species increased in relative abundance, both of which were exotic, whereas four species decreased in abundance (three native and one exotic). Native flies that decreased in relative abundance were all Syrphidae in the genus Helophilus.

@@

A recent survey highlighted that more than 90% of the papers studying pollinator responses to land use change focused solely on richness and abundance descriptors [@winfree2011]. But in addition to local (alpha) diversity and regional (gamma) diversity, we need to understand changes in turnover between sites (beta diversity). Environmental changes often results in a few species as “winners” and many species becoming “losers” [@Bartomeus2013b]. Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2018]. These changes can have important effects for pollination of native plant species and crops.

But museum specimen collection can provide much more information besides species occurrence records, and such information is important to capture when digitizing collections. This is specifically relevant to identify mechanisms of decline and adaptation. For example, one particularly important source of information is the date the specimen was collected, which allows the tracking of phenological advances congruent with contemporary climate change [#@Bartomeus2011]. In addition, museum labels on pollinator specimens often include information about the host plant on which the specimen was collected. This information is precious to understand past and present species interactions [#@Tylianakis2010ELE]. Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events [#@kleinj_raemakers]. Finally, museum specimens can be measured to track evolutionary changes by measuring specific specimen traits. This has been already explored to show tongue length [#@refAlaskaScience] and body size [#@Oliveira] shrinkage with climate and land use change.

#Conclucions

As illustrated in our examples, unleashing the power of museum collection data to answer pressing questions is at our hands, but requires the coordinated effort of many actors. Using two case studies, we show that collaboration with both museum curators and ecologists is key to adequately understanding the data and treating it appropriately. Researchers and curators should aim to make digitalized data readily available and easy to share. Centralization of regional and national museum data in existing global platforms such as GBIF would facilitate free and widespread access. However, specific datasets can also be stored in different webpages or database repositories (e.g. Universities and museum webpages, Dryad) as far as they are well documented and can be easely retrived and combined with other datasets using open science tools [e.g. ropensci, # @heltmint].

We also need to revolutionize the way that researchers collaborate with museums, which should start with fostering healthy bidirectional relationships. For instance, ecological researchers collect massive amounts of specimens, but these are often inappropriately vouchered (@Packer, @brief2), rendering them less useful for future research. To improve this, strong communication between museums and researchers is required, but this will only be achieved if there is not adequate funding, and recognition that accurate data recording and long-term preservation are critical for research [@WardTEE].
